Discovering Focused Microgenre Communities

### 2023-08-03 main.tex
Updates from Overleaf

\subsection{Heavy Metal}
Heavy Metal is one of the most storied macrogenre labels in the sociology of taste, as featured in the title of Bethany Bryson's \citeyearpar{bryson96} now classic paper. Since then, it has been the subject of scholarly debate regarding its possibly changing status in the cultural stratification order \citep{tampubolon2008revisiting, goldberg2011mapping, lizardo_skiles15}. Heavy Metal stood out in Bryson's study for several reasons. First, it was the most disliked genre in the 1993 General Social Survey culture module data---one of the most heavily analyzed data sources in the sociology of taste---hence the ``Anything But'' in the title of Bryson's eponymous piece. Second, it was the prototypical genre that was relatively more favored by working-class audiences but absolutely hated by individuals with high education. Thus, it served as the linchpin of Bryson's symbolic exclusion argument; omnivores were omnivorous up to a point, except when it came to those genres most favored by those with low education.

@@

Tampubolon's work indicates that just because Metal's audience is composed of low-education people does not mean that the rest of the less-educated population loves it; in fact, they hate it passionately. Tampubolon's analysis of ``Like'' responses also shows that Metal, rather than being exclusively preferred by a univorous group of ``Metalheads'' (who only prefer metal and dislike everything else), also makes it into the taste profile of one of two omnivorous groups uncovered in his analysis. Thus, Metal ends up straddling omnivores and univores. Because of this, even in the 1993 GSS data, its association with markers of status was ambiguous, leading Tampubolon to conclude that ``[p]reference or dislike for heavy metal is, therefore, orthogonal to status as measured by education'' \citeyearpar[][257]{tampubolon2008revisiting}, and that ``heavy metal is definitely not a low culture in the sense of culture `strongly associated' or very much liked by those with low education'' (\textit{ibid}).

@@

\subsubsection{Microgenre Analysis}
So, will the real heavy metal please stand up? Is Metal a form of emerging cultural capital with potential appeal to young middle-class people or a genre heavily favored by white men with low education? Or is that an outdated picture of the American musical field? Is Metal even correlated with status markers like education at all, as Tampubolon suggested? Is speaking of Metal as a ``youth'' music accurate, or has its audience been aging along with the genre (going on more than four decades and counting)? Given prior work, a strong expectation is that, at the very least, we should observe a microgenre partition separating the ``prototypical'' Metal (appealing to low-education white men) and perhaps a less-prototypical one, appealing to younger audiences with high-level cultural capital with other micro-variations, perhaps track generational distinction \citep{lizardo_skiles15}.

@@

\subsection{Salsa}
``it is plausible that increasing appreciation among high-status groups for a genre conventionally defined as ‘popular’ or ‘lowbrow’, will go hand in hand with its symbolic upgrading. And when a genre expands and spreads in terms of production, distribution and reception, it is likely to differentiate into distinct styles or subgenres. Under these conditions, it is to be expected that differential appreciations for different styles and forms within a given genre will develop, which may end up in a more or less clear hierarchy within this genre'' \citeyearpar[][62]{Bachmayer2014-pk}

``Taste in salsa music shows a strong internal, hierarchical order between ‘artistic’ versus ‘popular’ styles that is not only acknowledged and defined as such by experts and professionals within the field, but also structures the taste patterns of different social classes'' \citeyearpar[][62]{Bachmayer2014-pk}

\begin{figure}[ht!]
    \centering    
    \includegraphics[width=1.5\textwidth]{Tabs/reg-tab-latin.png}
    \caption{}
    \label{tab:reg-metal}
\end{figure}

@@

\section{Discussion}
Let us take stock. We began by discussing the network or ``relational'' revolution in the quantitative study of taste in sociology \citep{pachucki2010cultural}. We noted that many good things have come from considering survey data on taste using a relational lens, including a deeper specification of core concepts in the literature and even discovering new phenomena and empirical patterns. However, we also noted that survey data are as good as the labels chosen to collect the data. A resurgent line of critics questions whether the labels that appear in our most venerable survey-based studies are true ``genres'' in the sociological (or even stylistic) sense \citep{lena2015relational, vlegels2015music}. The labels are broad, likely to be interpreted by people in heterogeneous ways, and thus hide as much as they reveal. I noted recent attempts to just drop the idea of vague macrogenre labels and either study actual sociological genres on the ground (thus partially abandoning studies of audience segmentation or at least radically reconfigure them) or ``dropping the label'' \citep{sonnett2016ambivalence} by querying people about more focused objects of taste (e.g., performers within genres). These are all important and good developments, but we also noted that they may throw away a good thing. Upping the relationality by exploiting hidden patterns in the same old vague macrogenre-based data collected before can reveal focused microgenres.

I proposed to do this using recent developments in the discovery of \textit{overlapping communities} in networks \citep{ahn_etal10}, which partially answers two of the challenges of macrogenre critics: The fact that actual genres are overlapping and not crisply bounded. The fact that there is hidden heterogeneity within the broad labels we usually focus on. Our analysis contrasted the application of the usual techniques to the same data, conceived in the usual ``macro" way, and \textit{after} upping the relationality and discovering the microgenres hidden within. It is clear that focusing on microgenres reveals variation and patterns of cultural choice (as well as audience segmentation) that we would not have noticed using the standard approach (in this, the critics are right). Still, we could do this without dropping the label or querying people about hundreds of micro-styles (most of which they'd be unfamiliar with). Instead, we exploited the venerable principle, etched into classical approaches to defining genres in network terms \citep{dimaggio1987classification}, noting that if genres are defined by the people who choose them, then they are also defined by the other choices that people make when they choose them; the country that combines with one version of Classical may not be the same country that combines with a version of the Blues (and neither is the same as the country that refuses to combine with any other style).

>>>>

\subsubsection{Heavy Metal}
Heavy Metal is one of the most storied macrogenre labels in the sociology of taste, as featured in the title of Bethany Bryson's \citeyearpar{bryson96} now classic paper. Since then, it has been the subject of scholarly debate regarding its possibly changing status in the cultural stratification order \citep{tampubolon2008revisiting, goldberg2011mapping, lizardo_skiles15}. Heavy Metal stood out in Bryson's study for several reasons. First, it was the most disliked genre in the 1993 General Social Survey culture module data---one of the most heavily analyzed data sources in the sociology of taste---hence the ``Anything But'' in the title of Bryson's eponymous piece. Second, it was the prototypical genre that was relatively more favored by working-class audiences but absolutely hated by individuals with high education. Thus, it served as the linchpin of Bryson's symbolic exclusion argument; omnivores were omnivorous up to a point, except when it came to those genres most favored by those with low education.

@@

Tampubolon's work indicates that just because Metal's audience is composed of low-education people does not mean that the rest of the less-educated population loves it; they hate it passionately. Tampubolon's analysis of ``Like'' responses also shows that Metal, rather than being exclusively preferred by a univorous group of ``Metalheads'' (who only prefer metal and dislike everything else), also makes it into the taste profile of one of two omnivorous groups uncovered in his analysis. Thus, Metal ends up straddling omnivores and univores. Because of this, even in the 1993 GSS data, its association with markers of status was ambiguous, leading Tampubolon to conclude that ``[p]reference or dislike for heavy metal is, therefore, orthogonal to status as measured by education'' \citeyearpar[][257]{tampubolon2008revisiting}, and that ``heavy metal is definitely not a low culture in the sense of culture `strongly associated' or very much liked by those with low education'' (\textit{ibid}).

@@

\subsubsection{Metal Microgenre Analysis}
So, will the real heavy metal please stand up? Is Metal a form of emerging cultural capital with potential appeal to young middle-class people or a genre heavily favored by white men with low education? Or is that an outdated picture of the American musical field? Is Metal even correlated with status markers like education at all, as Tampubolon suggested? Is speaking of Metal as a ``youth'' music accurate, or has its audience been aging along with the genre (going on more than four decades and counting)? Given prior work, a strong expectation is that, at the very least, we should observe a microgenre partition separating the ``prototypical'' Metal (appealing to low-education white men) and perhaps a less-prototypical one, appealing to younger audiences with high-level cultural capital with other micro-variations, perhaps track generational distinction \citep{lizardo_skiles15}.

@@

\subsubsection{Salsa}
The Salsa music genre has not had as storied a career in the sociology of taste as Metal. In fact, it has been under-researched and, for the most part, ignored. I include it in the following analysis due to a shining exception to this pattern, represented in a high-quality recent study by \citep{Bachmayer2014-pk} which strongly suggests internal microgenre differentiation within this broad macrogenre label relevant to the reception and consumption side. These differentiations are keyed to hierarchical distinctions across Salsa microgenres based on social status and the concomitant internalization of class-based aesthetic dispositions \citep{lizardo2012reconceptualizing}.

Bachmayer et al. elicited a list of performers by asking experts to freely name performers and bands and then classify the music typical of the named acts into an ordinal rank based on their judgments of aesthetic quality (``superior,'', ``middlebrow,'' ``inferior''). They then selected the eight artists who were mentioned by at least three of the four expert informants and who displayed a high degree of concordance across experts in terms of their aesthetic rank. The eight compositions the experts judged to be typical of each artist's output were selected for guided interviews with forty first-generation Latinx immigrants in the Netherlands and Switzerland (there were no relevant cross-country differences). The most highly regarded expert pieces belonged to the sub-genre ``classic salsa,''---essentially early salsa performers that set the standard in the 1960s, 1970s, and 1980s---the middlebrow pieces were mainly variants of ``romantic salsa.'' The less aesthetically worthy pieces generally belonged to more recent styles of ``pop salsa styles.'' Note that this establishes a strict alignment between microgenre aesthetic standing and the historical sequencing of microgenre emergence within the salsa musicworld. We should thus expect this to be reflected in the generational location of each microgenre audience.

\begin{figure}[ht!]
    \centering    
    \includegraphics[width=1.5\textwidth]{Tabs/reg-tab-latin.png}
    \caption{}
    \label{tab:reg-latin}
\end{figure}

What did \citet{Bachmayer2014-pk} find? The short answer is that they found a microcosm of a Bourdieusian world---with some wrinkles. More educated respondents gravitated towards classic salsa and explicitly denigrated audiences that went for more popular styles---enacting a version of Ien Ang's \citeyearpar{ang2000dallas} ``ideology of mass culture'' in this field. High status respondents emphasized compositional and thematic complexity over ``fun'' aspects like rhythm and ``danceability,'' elements crucial in working-class Latinx immigrants' explanations for why they gravitated towards more romantic and pop styles (finding classic salsa sleepy and boring). Importantly, ``expert consumers'' (people who worked in the musical field in some capacity) demonstrated the same aestheticized appreciation for classic salsa and denigration of popular styles regardless of their class background. \citep[][62]{Bachmayer2014-pk} conclude that ``[t]aste in salsa music shows a strong internal, hierarchical order between `artistic' versus `popular' styles that \ldots structures the taste patterns of different social classes.''

This means that, at a minimum, we should expect the Salsa microgenres elicited by the link clustering procedure to reflect this hierarchical divide, with a microgenre more likely to be preferred by high-status respondents, particularly the higher-educated---the status marker that was most crucial in Bachmayer et al.'s study.

\subsubsection{Salsa Microgenre Analysis}
Table~\ref{tab:reg-latin} shows the result of a regression analysis analogous to that I specified for the Metal microgenres. The first column shows a model predicting choosing the ``Latin/Spanish/Salsa" macrogenre label using the usual socio-demographic covariates. The next three columns show the predictors of choosing each microgenre variation extracted from cutting the macrogenre label branch of the dendrogram at the required height. Looking at the first column, we would have concluded that choosing Salsa is not predicted by either respondent's ($p = 0.26$) or parent's ($0.17$) education, suggesting that taste for this genre is orthogonal to cultural capital and status. Instead, Salsa is mainly a genre preferred by young respondents (negative linear effect of age) who identify as Hispanic (no surprise there) and as women ($p < 0.05$).

@@

The microgenre analysis tells a different story. Looking at the third column of the table, predicting choice for the second of the induced microgenres, we find that, indeed, there is a micro-variation that is statistically more likely to be preferred by people of high education whose parents are also more likely to be highly educated ($p < 0.05$). Moreover, the age effect on the choice of this microgenre is \textit{reversed} compared to baseline expectations from the macrogenre regression. This microgenre is more likely to be preferred by older respondents (see Figure~\ref{fig:age-salsa}). In this way, this microgenre contrasts strongly with the first one extracted by the link clustering procedure, which is overwhelmingly more likely to be preferred by younger respondents. These results point to two conclusions: First, there is hierarchical differentiation among microgenres extracted from the ``Latin/Spanish/Salsa'' macrogenre, with one microgenre standing out by its exclusive appeal to high-cultural-capital respondents. Second, this hierarchical differentiation is keyed to age, consistent with the idea that the most prestigious Salsa microgenre is also the ``oldest'' and most traditional \citep{Bachmayer2014-pk}.

\section{Discussion}
Let us take stock. We began by discussing the network or ``relational'' revolution in the quantitative study of taste in sociology \citep{pachucki2010cultural}. We noted that many good things have come from considering survey data on taste using a relational lens, including a deeper specification of core concepts in the literature and even discovering new phenomena and empirical patterns. However, we also noted that survey data are as good as the labels chosen to collect the data. A resurgent line of critics question whether the labels that appear in our most venerable survey-based studies are true ``genres'' in the sociological (or even stylistic) sense \citep{lena2015relational, vlegels2015music}. The labels are broad, likely to be interpreted by people in heterogeneous ways, and thus hide as much as they reveal. I noted recent attempts to drop the idea of vague macrogenre labels and either study actual sociological genres on the ground (thus partially abandoning studies of audience segmentation or at least radically reconfigure them) or ``dropping the label'' \citep{sonnett2016ambivalence} by querying people about more focused objects of taste (e.g., performers within genres). These are all important and good developments, but we also noted that they may throw away a good thing. Upping the relationality by exploiting hidden patterns in the same old vague macrogenre data collected before can reveal focused microgenres.

I proposed to do this using recent developments in the discovery of \textit{overlapping communities} in networks \citep{ahn_etal10}, which partially answers two of the challenges of macrogenre critics: The fact that actual genres are overlapping and not crisply bounded. The fact that there is hidden heterogeneity within the broad labels we usually focus on. Our analysis contrasted the application of the usual techniques to the same data, conceived in the usual ``macro" way, and \textit{after} upping the relationality and discovering the microgenres hidden within. Focusing on microgenres reveals variations and patterns of cultural choice (as well as audience segmentation) that we would not have noticed using the standard approach (in this, the critics are right). Still, we could do this without dropping the label or querying people about hundreds of micro-styles (most of which they'd be unfamiliar with). Instead, we exploited the venerable principle, etched into classical approaches to defining genres in network terms \citep{dimaggio1987classification}, noting that if genres are defined by the people who choose them, then they are also defined by the other choices that people make when they choose them; the country that combines with one version of Classical may not be the same country that combines with a version of the Blues (and neither is the same as the country that refuses to combine with any other style).
